---
title: "The Influence of YouTube Videos on Song Popularity"
author: Bart van Vlerken
date: "Social Media and Web Analytics, 2021"
output: beamer_presentation
header-includes:
- \usepackage{booktabs}
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(rlist)
library(modelsummary)
```

## Motivation
* Rick Beato's 'What Makes This Song Great'
* Problems with Major Music Labels
* Fleetwood Mac's 'Dreams' Re-entering Charts
* Does the release of a ‘What Makes This Song Great?’ video on YouTube significantly increase the popularity metric for the song in question on YouTube, according to Google Trends?

## Data
* Google Trends
* Control Group
* Data

```{r, echo = FALSE, fig.align='center', out.width='60%'}
knitr::include_graphics('../output/figure_1.pdf')
```

## Methodology

* Difference-in-differences Approach

$$
Popularity_it = \beta_0 + \beta_1After_t + \beta_2Treated_i + \delta After_t * Treated_i + \varepsilon_it
$$

* Parallel Trends Assumption

```{r, echo = FALSE, fig.align='center', out.width='45%'}
knitr::include_graphics('../output/figure_2.pdf')
```

## Results
* delta = 6.8, se = 2.207, alpha = 0.002, 95% CI = {2.490, 11.146}

```{r, echo = FALSE, fig.align='center', out.width = '45%'}
knitr::include_graphics('../output/figure_3.pdf')
```

## Conclusion
* Popularity boost is likely due to (re)discovery of music
* Main Takeaway for Major Labels